Effective Inclusion of Electronic Polarization to a Negatively Charged Membrane

### 2019-03-08 Manuscript/manuscript.tex
Update text

\begin{figure}[tb!] 
  \centering 
  \includegraphics[width=\figwidth]{../img/ecc_pops/Order-parameters_form-factors_exp-L17-ECC-lipids.pdf}
  \includegraphics[width=\figwidth]{../img/ecc_pops/pops_chemfig.pdf} 
\hfill
  \caption{\label{simVSexpNOions_POPS} 
    \textbf{Top:} X-ray scattering form factors from simulations with Lipid17/Dang \citep{lipid17-future, dang2006} and 
    the ECC-POPS/ECC-ions \cite{martinek17, Pluhackova2016} compared with experiments~\citep{kucerka14} at 298~K. 
    \textbf{Middle:} Order parameters of POPS head group, glycerol backbone and acyl chains  
    from simulations with the Lipid17 \citep{lipid17-future} and the ECC-POPS models 
    compared with experiments at 298~K. \citep{nmrlipids_proj4}
    Open/closed symbols are used for palmitoyl/oleoyl chains of POPS. 
    \textbf{Bottom:} The chemical structure of POPS and the labeling of the carbon segments. 
  }  
\todo{J: flip the chemical structure so that it agrees with the left-to-right reading of the plot.}
\end{figure}

\begin{table}[tb!] 
\centering
  \caption{Values of the area per lipid (APL) of POPS bilayers 
	   at 298~K with only \ce{Na^+} counterions and no additional ions. 
           Standard deviations are given. \label{tab:apls} } 
  \begin{tabular}{l|c } 
    \multicolumn{2}{c}{POPS} \\
    model          & APL / Å$^2$    \\ 
    \hline 
    Lipid17/Dang              & 53.5$\pm$ 1.7   \\ 
    Lipid17/ff99              & 57.9$\pm$ 1.7   \\ 
    \hline 
    ECC-POPS/ECC-ions         & 59.8$\pm$ 1.8   \\ 
    \hline 
    experiment (SDP model) \citep{kucerka14} & 62.3  \\ 
    \hline 
  \end{tabular} 
\end{table}

We compared X-ray scattering form factors and NMR order parameters of bilayers
in pure water without any ions (or only counter ions)
from simulations and experiments. 
The experimental X-ray scattering form factors 
of a bilayer are well reproduced by ECC-POPS, 
while for Lipid17/Dang they are in a slight disagreement
(see Fig.~\ref{simVSexpNOions_POPS}). 
The agreement of Lipid17 can be improved by using different parametes for ions \cite{aqvist90},
which, however, bring other kinds of artifacts with them 
(e.g., incorrect pairing of ions, condensation at larger salt concentrations \cite{kohagen16, chen07, nmrlipids_proj4}).

%The area per lipid is often used as a relatively simple structural parameter reporting on the bilayer properties and the packing of lipids. 
In order to obtain the area per lipid from experiments, modeling is used on top of the scattering factors \citep{kucerka14}. 
In ocntrast, this property is easily extracted directly from simulations with flat phospholipid bilayers. 
We compare the values from experiments and simulations in Table~\ref{tab:apls}.

@@

While the agreement between the scattering form factors 
from the simulation of a pure POPS bilayer and experiment 
are excellent (Fig.~\ref{simVSexpNOions_POPS}),
there is a non-negligible difference between the values of the area per lipid in Table~\ref{tab:apls}. 
Since both values are derived from the scattering form factors through modeling of the electron density of the bilayer,
we cannot decide, which of the values is more reliable. 
In general, we can conclude that ECC-lipids
%the presented lipid models with ECC-lipids 
reproduce the experimental structural parameters of the lipid bilayers 
with a comparable accuracy to existing state-of-the-art lipid models~\citep{botan15, ollila16, Pluhackova2016}.

@@

The head group and acyl chain order parameters within ECC-lipids
are in general in a good agreement with the experimental values 
as shown in Fig. \ref{simVSexpNOions_POPS}. 
The acyl chain order parameters in particular agree with the experimental data within error bars.
The order parameters of the head groups are at an accuracy comparable to 
other currently available classical models of lipids \citep{botan15, catte16, Pluhackova2016, nmrlipids_proj4}. 
In particular, ECC-POPS is closer to the experimental order parameters 
than Lipid17 with either Dang \cite{smith94,chang1999,dang2006} 
or ff99 ions \citet{aqvist90}.

%The head group order parameters $\alpha$ and $\beta$ are highly relevant for this work,
%as they are being used in the electrometer concept \cite{altenbach84, catte16, melcr18}.
%For POPC in pure water, the order parameter $\beta$ agrees well with the experiment, while the order parameter $\alpha$ is somewhat lower. 
The head group order parameters of PS are a bit more complicated 
compared to PC as the order parameter $\alpha$ exhibits a notable forking 
(i.e., the hydrogen atoms in the group give rise to different order parameters, see Fig.~\ref{simVSexpNOions_POPS}).
One of the order parameters of ECC-POPS, $\alpha_1$, agrees well with the experiment, 
while the other, $\alpha_2$, adopts a higher value underestimating the experimentally reported forking. 
There is only one order parameter $\beta$ in POPS, 
which has a higher value closer to zero in the ECC-lipids model than in experiment. 
Such a feature suggests that the model likely overestimates the orientational freedom of the PS head group.

@@

It has been pointed out in previous experimental works
that the conformational flexibility of PS head groups 
is different from other lipids \cite{browning80}. 
While the static shielding tensor of PS is not substantially different from PC or PE, 
the $^{31}$P~$T_1$ relaxation times are significantly shorter for PS. 
It is suggested that the PS head group is more rigid compared to the other phospholipids, 
which is also corroborated by a larger chemical shielding anisotropy of PS \cite{browning80}.

We have investigated this effect using the mean orientations of P-N and \ce{COO^-} vectors 
(defined and plotted in Fig.~\ref{fig:delta_ordPar_CaCl_PCPS}). 
While the mean angle between the P-N vector of PC head group and the membrane normal
decreased by 11$^\circ$ from 64$^\circ$ to 55$^\circ$
 at 620~mM buffer concentration of \ce{CaCl2},
the mean orientation of the P-N vector in the PS head group has 
decreased only by 3$^\circ$ from 67$^\circ$ to 64$^\circ$ 
in the same concentration range.

Such a largely diminished flexibility 
may arise from the binding of calcium to the \ce{COO^-} group in PS. 
Whe have measured the mean orientation of the \ce{COO^-} group 
as the connector of the carbon atoms 
forming the bond between the group and the $\beta$-carbon of the phospholipid. 
The mean angle between the \ce{COO^-} vector and the membrane normal
changes by $11^\circ$ from $62^\circ$ to $73^\circ$ (more towards the membrane interior)
 at 620~mM buffer concentration of \ce{CaCl2}.
The conformational flexibility of the PS head group is hence restrained 
by the binding of calcium cations to the \ce{COO^-} group, 
which reorients towards the membrane interior and to the phosphate region, 
where calcium cations bind predominantly.

@@

The binding of calcium cations to the phospholipid bilayer is shown 
in the density profiles of \ce{Ca^{2+}}, \ce{Na^+} counterions 
and also \ce{Cl^-} ions in Fig.~\ref{fig:cacl-dens_PCPS}. 
Clearly, after accounting for the electronic polarizability, 
the population of the cations at the membrane interface is significantly diminished. 
\todo{JOE: The density plot still needs to be changed to show both Lipid17 and ECC-lipids, and to support this argument.}
Still, the dominant contribution to the binding of calcium comes from the phosphate groups. 
Interestingly enough,
although the \ce{COO^-} groups increase the affinity of calcium towards PS compared to PC,
 there is no extra density at the outmost surface of the membrane. 
This can be understood from the point of the structural rearangements of the \ce{COO^-} group, 
which take place during the interactions with calcium cations
described above and in Fig.~\ref{fig:delta_ordPar_CaCl_PCPS}.

\todo{JOE: suggestion -- what is the charge at the membrane interface of memb+counterions -vs- memb+CaCl2 concnetration?
                         this could relate well to the zeta potential measurements and the Fig 1 in \cite{melcrova16}. }

\begin{table}[tb!] 
\centering
  \caption{Bulk concentrations, $C _{Ca}$, and buffer concentrations, $C' _{Ca}$, of Ca$^{2+}$;
           relative surface excess of calcium with respect to water ($\Gamma_{Ca}^{\rm water}$); 
           and percentages of the population 
           of bound Ca$^{2+}$ to various moieties 
           in a neutral membrane composed of POPC
           and in a negatively charged membrane with the composition 5\,PC:1\,PS.
           \label{tab:binding}} 
  \begin{tabular}{ l | c c } 
	                     &  5\,POPC:1\,POPS &  POPC   \\
	\hline
	$C _{Ca}\,/\,\mathrm{mM}$  &  $240\pm 10 $  &  $280\pm 10 $  \\
	$C'_{Ca}\,/\,\mathrm{mM}$  &  $400\pm 10 $  &  $350\pm 10 $  \\
	$\Gamma_{Ca}^{\rm water}\, / \,\mathrm{nm}^{-2}$  &  $0.24 \pm 0.01 $  &  $0.06 \pm 0.01 $  \\
	\hline
                             &  \multicolumn{2}{c}{ } \\
        interacting moiety   &  \multicolumn{2}{c}{percentage of bound \ce{Ca^{2+}} } \\
	\hline
	     PC              &   59   &  100   \\
	     PO$_4$    in PC &   41   &   67   \\
	     carbonyls in PC &   <1   &   ~1   \\
	\hline
	     PS              &    8   &        \\ 
	     PO$_4$  in PS   &    2   &        \\
	     COO$^-$ in PS   &    4   &        \\
	     carbonyls in PS &   <1   &        \\
	\hline
	both PC and PS       &   33   &        \\
  \end{tabular} 
\end{table}

% Gamma and higher affinity of PS to CA2+
The higher affinity of calcium to PS head groups compared to PC
is well demonstrated using the relative surface excess, $\Gamma ^w _{Ca}$,
summarized in Table~\ref{tab:binding}. 
While $\Gamma^{w}_{Ca}$ at a neutral POPC bilayer
is       $\Gamma^{w}_{Ca} = 0.06\mathrm{nm^{-2}}$ \cite{melcr18}, 
its value for the PC:PS (5:1) bilayer is $\Gamma^{w}_{Ca} = 0.24\mathrm{nm^{-2}}$
at comparable buffer concentrations  $350 \, \mathrm{mM}$ resp. $400\, \mathrm{mM}$.

% Description of no. conctacts counting
Such a significantly increased binding also has an impact 
on the structure and populations of the bound calcium cations. 
We have counted contacts between the cations and the oxygen atoms of the lipids
similarly as was done in \cite{melcr18} 
to obtain such a detailed information 
about the interactions of \ce{Ca^{2+}} with various moieties in POPC or POPS. 
The threshold for counting a contact was set to $0.3\,\mathrm{nm}$, 
which encompasses the first peak of the radial distribution function between the cations and the oxygen atoms of the lipids.

% Populations from contacts
The percentages of the populations of membrane-bound calcium cations for various membrane moieties 
are summarized in Table~\ref{tab:binding}.
Even though the negatively charged membrane contains only 18\% of POPS, 
on average 41\% of the total population of bound calcium cations is in contact with PS lipids
with 8\% bound only to them. 
% Where Calcium interacts
While POPC interacts with the calcium cations almost entirely through its phosphate group 
in both neutral \cite{melcr18} and negatively charged membranes, 
the inteaction with POPS lipids happens more through the \ce{COO^-} group than through the \ce{PO4} group. 
Interactions of \ce{Ca^{2+}} with carbonyl groups are also present for both POPC and POPS,
however, they are always accompanied by interactions with phosphate groups. 
This is in contrast to 
simulations with the Berger force field \cite{berger97,mukhopadhyay04}
combined with fluorescent and vibrational sum frequency spectroscopy,
which suggest a significant
calcium binding also to the carbonyls groups \cite{melcrova16}. 
This feature arises likely from the inaccuracy of the Berger model 
and its overestimated binding of cations to that region \cite{nmrlipids_proj4}.

Recent simulations reveal similar trends for the populations at various PS moieties. \cite{valentine18, hallock18}
The results from infrared spectroscopy combined with 
simulations with the CHARMM36 model 
including a correction for overestimated binding of calcium \cite{valentine18}
suggest that POPS interacts mostly through it \ce{COO^-} group. 
In contrast, the combination of NMR experiments 
and MD simulations with the same model but without the correction 
also reveal long-lived interactions of \ce{Ca^{2+}} with \ce{PO4} groups. \cite{hallock18}
The content of POPS in the bilayers in that work is, however, 
much higher above the threshold after which 
\ce{Ca^{2+}} cations induce lipid clustering. \cite{hauser77,kurland79,hauser85,feigenson86,mattai89,roux90,roux91}
At such highly charged membranes,
POPS lipids are found to form even four-membered clusters \cite{hallock18}. 
Our simulations are, however, at different conditions,
in which the clustering of POPS does not happen, 
and PS lipids prefer 1:1 ratio with \ce{Ca^{2+}}
as plotted in Fig.~\ref{fig:cacl_complexes}
showing relative probabilities of \ce{Ca^{2+}} complexes with a certain number of PC or PS lipids. 
Compared to a neutral POPC bilayer
the negatively charged bilayer has its stoichiometry 
shifted towards complexes with three phospholipids and one calcium. 
This is obviously due to the presence of POPS, 
which also contributes to the pre-formed one- and two-membered clusters of POPC. 
Clusters of four or more lipids were not observed in either membrane.

THE REMAINING PIECES TO BE DISCUSSED:

% Ca2+ induces much slower dynamics of lipids when PS is present,
% this does not happen with PC
% uses NBFIX
From Ref.~\cite{valentine18}://
To summarize, calcium binding to PS lipids slows the dynamics at the ester position and introduces more heterogeneous molecular environments. 
The slowdown cannot be attributed to complete ester dehydration or decreased hydrogen bonding to water, so we interpret these results as evidence that the lipid headgroups adopt more rigid but disordered conformations on the molecular level. 
The decrease in dynamics observed experimentally indicates that the network of hydrogen bonds between lipids and water at the polar-nonpolar interface becomes slower, as it was shown by Karathanou and Bondar (34). 
This is bolstered by recent NMR work showing that binding to Ca2+ induces two different rigid conformations in the glycerol backbone of lipid headgroups (18).
We have investigated the molecular origin of the rigidification and decreased lateral diffusion, finding that calcium binding does not dehydrate the ester region, but even with only 10\% PS lipids, the local dynamics slow around the esters. The slower local dynamics provide a molecular explanation for the Ca2+-dependent decreases in fluidity observed on longer length and timescales.
There is considerable room for further improvement in the force fields for ion-membrane interactions. Particularly, small differences in short-range interactions may determine the local geometries (34
). Most of our conclusions are, however, based on long-range effects (i.e., electric field generated by the ion at significant distances from the carbonyl groups), which are less sensitive to the details of the force field.

Their Figure 5 \cite{hallock18} shows the distributions of O-Ca-Cb-N dihedral angle
revealing separate populations of the PS head groups. 
These two structures are also compared in therm of P-Cg distances, 
which can be measured from NMR REDOR exp. 
It is concluded that since the two distinct populations in P-Cg distance can be represented with the 
dihedral distributions, 
they also likely represent the distinct populations. 
 -> plotting the distribution of O-Ca-Cb-N dihedral angle may be also a good idea for us (SI?)
     -> NMRlipids 4 in SI \cite{nmrlipids_proj4} plots this information: 
           - C36 agress with the distribution in \cite{hallock18}
           - Lipid17 (and Slipids) are almost identical to C36
           - but the structure of the COO- is highly different (dih angle Ca-Cb-Cg-Og)

@@

@@

@@

\section{SI}

>>>>

The form factors from scattering and C-H bond order parameters from NMR experiments are good measures to validate
the lipid bilayer structure in MD simulations against experiments because they
are related to the bilayer dimensions (area per molecule and thickness) and
conformational fluctuations of individual lipids, respectively, and 
can be directly calculated from MD simulations~\cite{ollila16}.
The X-ray scattering form factors and C-H bond order parameters in
the headgroup and glycerol backbone region of POPS lipid bilayer are already available in the literature \cite{pan14,nmrlipids_proj4}.
Here, we extend the available experimental dataset of POPS lipids by measuring the C--H bond
order parameters also from the acyl chain region using the R-PDLF experiments as described previously \cite{ferreira13}.
We first determined the order parameters of each peak corresponding different C--H bonds detected
in the first dimension (Fig. \ref{R-PDLF}) from the dipolar slices in the indirect dimension (Fig. \ref{R-PDLFslices}).
Because the exact assignment of these peaks was not possible, we assigned the
acyl chain order parameters to mimic the profile predicted by the MD simulations (Fig. \ref{simVSexpNOions_POPS}).
\todo{Tiago has to check and update this and figures in SI.}
\begin{figure}[tb!] 
  \centering 
  \includegraphics[width=\figwidth]{../img/ecc_pops/Order-parameters_form-factors_exp-L17-ECC-lipids.pdf}
  \includegraphics[width=\figwidth]{../img/ecc_pops/pops_chemfig.pdf} 
\hfill
  \caption{\label{simVSexpNOions_POPS} 
    \textbf{Top:} X-ray scattering form factors from simulations with Lipid17/Dang \citep{lipid17-future, dang2006} and 
    the ECC-POPS/ECC-ions \cite{martinek17, Pluhackova2016} compared with experiments~\citep{kucerka14} at 298~K. 
    \textbf{Middle:} Order parameters of POPS head group, glycerol backbone and acyl chains  
    from simulations with the Lipid17 \citep{lipid17-future} and the ECC-POPS models 
    compared with experiments at 298~K. \citep{nmrlipids_proj4}
    Open/closed symbols are used for palmitoyl/oleoyl chains of POPS. 
    \textbf{Bottom:} The chemical structure of POPS and the labeling of the carbon segments. 
  }  
  \todo{J: flip the chemical structure so that it agrees with the left-to-right reading of the plot.} \\
  \todo{The experimental order parameters from sn-1 chain are missing from the figure.} \\
  \todo{We should use the same color coding for the models as in ECC-POPC paper, now it is opposite.} \\
  \todo{Inverted axis and positive value as label would be more clear for the order parameter figure.}
\end{figure}

To evaluate the effect of ECC on the structural properties of POPS lipid bilayer,
we compared the X-ray scattering form factors, C--H bond order parameters and area per molecules
between simulation with original Lipid17 \cite{??} combined with the Dang ion parameters \cite{??},
ECC-POPS simulation and experiments (Fig. \ref{simVSexpNOions_POPS} and table \ref{tab:apls}).
The larger acyl chain order parameters, deviation in the form factor and smaller area per molecule 
in the Lipid17 simulation with Dang ions when compared with experiments indicates that this simulation
predicts too dense lipid bilayer. In previous study \cite{nmrlipids_proj4}, Lipid17 simulated with {\AA}qvist ions~\cite{aqvist90}
gave larger area per molecule (57 \AA$^2$), probably due to the weaker counterion binding affinity.
The Dang ions are used to here to avoid the artificial pairing of ions and condensation at larger salt
concentrations observed with {\AA}qvist \cite{kohagen16, chen07, nmrlipids_proj4} when analysing the effect of ECC. 
The scattering form factors and acyl chain order parameters from ECC-POPS are in good agreement with
experiments (see Fig.~\ref{simVSexpNOions_POPS}), indicating the the bilayer dimensions and acyl chain
conformations are well described by the force field. The area per lipid from ECC-POPS simulation is slightly
smaller than the value reported from SDP model \cite{kucerka14} (table \ref{tab:apls}), but the values
are within each others error bars. 
%In order to obtain the area per lipid from experiments, modeling is used on top of the scattering factors \citep{kucerka14}. 
%In ocntrast, this property is easily extracted directly from simulations with flat phospholipid bilayers. 
%We compare the values from experiments and simulations in Table~\ref{tab:apls}. 
%While the agreement between the scattering form factors from the simulation of a pure POPS bilayer and experiment 
%are excellent (Fig.~\ref{simVSexpNOions_POPS}), there is a non-negligible difference between the values of the area per lipid in Table~\ref{tab:apls}. 
%Since both values are derived from the scattering form factors through modeling of the electron density of the bilayer,
%we cannot decide, which of the values is more reliable.

\begin{table}[tb!] 
\centering
  \caption{Values of the area per lipid (APL) of POPS bilayers 
	   at 298~K with only \ce{Na^+} counterions and no additional ions. 
           Standard deviations are given. \label{tab:apls} } 
  \begin{tabular}{l|c } 
    \multicolumn{2}{c}{POPS} \\
    model          & APL / Å$^2$    \\ 
    \hline 
    Lipid17/Dang              & 53.5$\pm$ 1.7   \\ 
%    Lipid17/ff99              & 57.9$\pm$ 1.7   \\ 
%    \hline 
    ECC-POPS/ECC-ions         & 59.8$\pm$ 1.8   \\ 
%    \hline 
    experiment (SDP model) \citep{kucerka14} & 62.7$\pm$ 1.3  \\ 
%    \hline 
  \end{tabular} \\
  \todo{I removed the value from Lipid17/ff99 simulation because this was not listed in Table I
    and I think that it is more clear if we refer only to the NMRlipids IV result in the text.
    I also changed the experimental value to correspond the value in the cited paper and added the error bar.}
\end{table}

@@

@@

Despite the success in simulating bilayer dimensions and acyl chain conformations,
lipid force fields have typically problems in capturing the correct glycerol backbone and
headgroup order parameters and conformations \cite{botan15,ollila16,nmrlipids_proj4}.
This is the case also for Lipid17 POPS simulations, where the headgroup and glycerol backbone
order parameters are quite far from experimental values with both Dang (Fig. \ref{simVSexpNOions_POPS})
and {\AA}qvist \cite{nmrlipids_proj4} ions. The values from ECC-POPS simulation are closer,
but not in full agreement with experiments (Fig. \ref{simVSexpNOions_POPS}).
Especially, the experimentally observed more negative values of the $\beta$-carbon
and the larger forking (difference of order parameters between hydrogens attached to the
same carbon) of the $\alpha$-carbon in POPS than in POPC \cite{nmrlipids_proj4}
are not reproduced by neither of the models here. These differences are previously related
to a more rigid structure of PS headgroup and the stiffness of the C$_\alpha$-C$_\beta$-C$_\gamma$-O$_\gamma$
dihedral, which is \todo{the dihedral should be plotted and sentence finished.}

%The head group order parameters $\alpha$ and $\beta$ are highly relevant for this work,
%as they are being used in the electrometer concept \cite{altenbach84, catte16, melcr18}.
%For POPC in pure water, the order parameter $\beta$ agrees well with the experiment, while the order parameter $\alpha$ is somewhat lower.

In conclusion, the ECC-POPS model reproduces the lipid bilayer dimensions and
acyl chain conformations in good agreement with experiments, but has room for
improvement in the glycerol backbone and headgroup region. Therefore,
its structural accuracy is in the same level as in other state-of-the-art
lipid force fields \cite{botan15, ollila16, Pluhackova2016, nmrlipids_proj4}.

@@

@@

%%% Present the density profiles (different concentrations for both POPC and mixed)
The increase in the amount of bound calcium cations from pure POPC \cite{melcr18}
to the mixed negatively charged bilayer containing POPS
is well demonstrated using the relative surface excess, $\Gamma ^w _{Ca}$,
summarized in Table~\ref{tab:binding}. 
Distributions of \ce{Ca^{2+}}, \ce{Na^+} counterions and also \ce{Cl^-} 
are plotted in Fig.~\ref{fig:cacl-dens_PCPS} for the negatively charged bilayer. 
In contrast to \ce{KCl} or added concentrations of \ce{NaCl}, 
\ce{Na^+} counterions are substituted with \ce{Ca^{2+}} even at low concentrations of \ce{CaCl2}. 
The increasing concentration of \ce{CaCl2} and, hence, a higher amount of bound \ce{Ca^{2+}}
also attracts \ce{Cl^-} anions to the bilayer 
as can be seen from its growing density at the interface.

The relative surface excess of calcium, $\Gamma^{w}_{Ca}$, also reflects
the higher affinity of the negatively charged PC:PS (5:1) membrane compared to a neutral POPC bilayer; 
its value for the latter is       $\Gamma^{w}_{Ca} = 0.06\mathrm{nm^{-2}}$, 
while for the PC:PS mixture it is $\Gamma^{w}_{Ca} = 0.24\mathrm{nm^{-2}}$
at comparable concentrations $C _{Ca} = 260\pm 30 \mathrm{mM} $.

\todo{SAMULI: This paragraph in unclear. Should be clarified and the changes
  of headgroup orientations in PC and PS clearly discussed.
}
%%% discuss the table with Gammas and follow up with molecular details, where do cations bind? Support with the spatial density figure
The density profiles of ions suggest that
the dominant contribution to the binding of \ce{Ca^{2+}} to phospholipid bilayers
comes from the interactions with the phosphate groups in both POPC and POPS. 
%The population analysis also suggests 
%that the calcium cations prefer to reside in the phosphate region of the membrane. 
%Such a finding is further validated with a more detailed analysis using Markov state modeling (MSM) \citep{Pande_MSM_paper} \todoi{Add papers reviewing MSM, e.g. recent Pande's paper.}. 
%The set of states of a calcium cation 
%encompassed all possible combinations of up to three surrounding lipids. 
%In the case of PC, we used only the phosphate moitety, which forms the dominant contribution for calcium binding.
%For PS we also distinguished configurations in which calcium interacts with the carboxylate moiety. 
%All possible combinations of such states were used to build a Markov model at a lag time $25\,\mathrm{ns}$ using pyEMMA code by \citet{pyemma} \todoi{add citation for pyemma}. 
%The resulting MSM was validated using Chapman-Kolmogorov test \citep{FrankNoe_papers_MSM} \todoi{Add citation for Noe's papers on MSM, especially Chapman Kolmogorov test.}. 
%The stationary distribution of the states reveals a strong preference of the calcium cations to reside in the phosphate region of the mixed PC-PS membrane. 
%When interacting with PS, configurations with the phosphate moiety from either PC or PS dominate the population.
%Moreover, states containing interactions with the carboxylate group in PS 
%bear higher probability when interacting also with the phosphate group of the same lipid or other lipids. 
This is also reflected in the shift of the mean orientation 
of the \ce{COO^-} group from $62^\circ$ to $73^\circ$ (420~mM \ce{CaCl2})
which was measured as the connector of the carbon atoms 
forming the bond between the group and the $\beta$-carbon of the phospholipid. 
The interactions of the carboxylate group in PS with calcium and other phosphate groups
shed light into the qualitatively different response of the head group order parameters $\alpha$ and $\beta$ in PS compared to PC 
(see Figs.~\ref{fig:delta_ordPar_CaCl} and~\ref{fig:delta_ordPar_CaCl_PCPS}). 
We find that the complex response of the head group order parameters of POPS 
is affected by the conformational changes of the carboxylate group,
which is attracted more towards the phosphate region, 
where the calcium cations dominantly bind. 
In line with the experimental work by \citet{browning80},
this is also very likely the reason, 
why the magnitude of the P-N vector change in POPS is diminished compared to POPC, 
which is not restrained by an additional cation binding group like \ce{COO^-} in POPS.

Further details about the interactions of \ce{Ca^{2+}} with various moieties in POPC or POPS
were obtained by counting contacts between the cations and the oxygen atoms of the lipids
similarly as was done in \cite{melcr18}. 
The threshold for counting a contact was set to $0.3\,\mathrm{nm}$, 
which encompasses the first peak of the radial distribution function between the cations and the oxygen atoms of the lipids.

\begin{table}[tb!] 
\centering
  \caption{Bulk concentrations, $C _{Ca}$, and molar fractions, $C' _{Ca}$, of Ca$^{2+}$;
           relative surface excess of calcium with respect to water ($\Gamma_{Ca}^{\rm water}$); 
           and percentages of the population 
           of bound Ca$^{2+}$ to various moieties 
           in a neutral membrane composed of POPC
           and in a negatively charged membrane with the composition 5\,PC:1\,PS.
           \label{tab:binding}} 
  \begin{tabular}{ l | c c } 
	                     &  5\,POPC:1\,POPS &  POPC   \\
	\hline
	$C _{Ca}\,/\,\mathrm{mM}$  &  $240\pm 10 $  &  $280\pm 10 $  \\
	$C'_{Ca}\,/\,\mathrm{mM}$  &  $400\pm 10 $  &  $350\pm 10 $  \\
	$\Gamma_{Ca}^{\rm water}\, / \,\mathrm{nm}^{-2}$  &  $0.24 \pm 0.01 $  &  $0.06 \pm 0.01 $  \\
	\hline
                             &  \multicolumn{2}{c}{ } \\
        interacting moiety   &  \multicolumn{2}{c}{percentage of bound \ce{Ca^{2+}} } \\
	\hline
	     PC              &   59   &  100   \\
	     PO$_4$    in PC &   41   &   67   \\
	     carbonyls in PC &   <1   &   ~1   \\
	\hline
	     PS              &    8   &        \\ 
	     PO$_4$  in PS   &    2   &        \\
	     COO$^-$ in PS   &    4   &        \\
	     carbonyls in PS &   <1   &        \\
	\hline
	both PC and PS       &   33   &        \\
  \end{tabular} 
\end{table}

The percentages of the populations of membrane-bound calcium cations for various membrane moieties 
are summarized in Table~\ref{tab:binding}.
Even though the negatively charged membrane contains only 18\% of POPS, 
approximately 41\% of the total population of bound calcium cations is in contact with PS lipids
with 8\% bound only to them. 
This corroborates the intrinsically higher affinity of PS lipids to calcium cations compared to neutral PC lipids. 
POPC interacts with the calcium cations almost entirely through its phosphate group 
in both neutral and negatively charged membranes. 
Interactions of \ce{Ca^{2+}} with carbonyl groups are also present, 
however, they are always accompanied by interactions with phosphate groups. 
%but a very small negligable amount much below 1\%. 
\todo{Also COO- vs. phosphate binding should be discussed and the results should be
  discussed with respect to the literature: https://doi.org/10.1016/j.bpj.2018.08.044 and
  Melcrova et al. Scientific Reports | 6:38035 | DOI: 10.1038/srep38035 }
%
\todo{EDIT: Taken from NMRlipids4 as an interesting information, to be edited.}
The dehydrated complexes of PS headgroup and calcium ions can also lead to the
phase separation \cite{hauser77,kurland79,hauser85,feigenson86,mattai89,roux90,roux91}.
%
CHARMM36 force field \cite{klauda10,venable13} combined with 2D infrared spectroscopy
suggests that calcium ions interacts only with the carboxylate group of PS lipids \cite{valentine18}, while
the same force field without the NBfix parameters together with the NMR chemical shifts and
REDOR experiments suggests a significant binding affinity also to the phosphate region \cite{hallock18}.
On the other hand, simulations with the Berger force field \cite{berger97,mukhopadhyay04}
combined with fluorescent and vibrational sum frequency spectroscopy suggested a significant
calcium binding also to the carbonyls in the acyl chains \cite{melcrova16}.

@@

Relative probabilities of \ce{Ca^{2+}} complexes with a certain number of lipids are presented in Fig.~\ref{fig:cacl_complexes}. 
Calcium cations that are bound only to PC in the mixed bilayer with PS 
behave similarly as in the pure PC bilayer
maintaining comparable probabilities for clustering one, two, or even three PC lipids together. 
In contrast, PS lipids prefer 1:1 ratio with \ce{Ca^{2+}},
which may also be due to their low molar fraction in the the mixed bilayer. 
In total, however, the negatively charged membrane has its stoichiometry 
shifted towards complexes with three phospholipids and one calcium. 
This is obviously due to the presence of POPS, 
which also contributes to the pre-formed one- and two-membered clusters of POPC. 
Clusters of four or more lipids were not observed in either membrane.

%This is also reflected in the increased probability 
%of a single phospholipid interacting with two \ce{Ca^{2+}} cations 
%compared to the neutral POPC bilayer, 
%where such a probability is almost negligable. 
% Not shown through numbers, however, I have the plots to back it.

@@

%In addition to such estimates of the time scales, we used the Markov model on top of the simulation with the negatviely charged mixed bilayer
%to calculate the time of the mean first passage of calcium from solution to the membrane (and in reverse) resulting in $55\,\mathrm{ns}$  ($165\,\mathrm{ns}$).

%\todo{Fluxes: committor analysis, dominant fluxes (table and figure)}
%From the spectrum of the transition matrix, we also observe that 
%the slowest transitions are associated with the binding to two or three phosphate moieties from either PC or PS,
%which are also among the states with the highest probabilities.
%The net fluxes of calcium cations from solution to such states also form a large contribution to the total flux ($\approx 45\%$). 
%\todo{Make a table and a figure of the state probabilities and fluxes to support this statement.}. 
%Analysis of the possible binding pathways of calcium cations to the negatively charged mixed bilayer 
%reveal that the cations mostly enter the membrane bound states directly from solution 
%without complicated transitions at the time scales of the order of the lag time of the Markov state model, $25\,\mathrm{ns}$.

Provide density plots and show, where and how the cations bind to the membrane.

@@
